Wood density and the growth-dependent & independent mortality

### 2015-04-09 ms.Rmd
Remove old results, add notes on cross-val

##Introduction
Understanding who lives or dies, and why, is critical to accurately predict population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr]. In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more likely to die [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr]. Empirical studies have broadly supported this assumption, with many studies showing that the probability of dying decreasing exponentially with growth rate [Fig 1; @Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw], and consequently most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db] and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use this growth-mortality relationship as the predominant, and in many cases, only cause of plant mortality [@Hawkes:2000ib; @McDowell:2011dr].

@@

##Data

@@

##Model Description

@@

@@

##Results
**Take home message:**
Wood density decreases intercept (log_a) & further increases the steepness of the growth-mortality slope (log_b). It also decreases the growth-independent hazard log_ho

```{r Model output, echo=FALSE}
print(stan_output_combined,probs = c(0.025,0.975), digits_summary = 3)
```

```{r Coefficient plot, fig.height=6,fig.width=6, echo=FALSE}
coeff_plot(stan_output_combined, params=c('c1','c0_mu','b1','b0_mu','a1','a0_mu'),
           labels = c('ho x wd','ho','beta x wd','beta','alpha x wd','alpha'),
           xlab = 'cloglog effect size', transform = NULL)
```

**Fig 7** log centered and scaled coefficients ± 95 (thin line) and 80% (thick line) Bayesian Credible Intervals. log_a = intercept; log_b = slope, log_ho = asymptote.

```{r Hazard rates,fig.height=6,fig.width=6, echo=FALSE}
par(mfrow=c(2,1),mar=c(4,3,1,1), xaxs='i')
plot_mortality(model_fit = stan_output_combined ,model_data = stan_data, cov_name = 'rho',
               wd_values = c(200,800), growth_name = 'growth_dt', mortality_curve = FALSE, xaxis_on = FALSE, haz_lim=c(0,2))

plot_mortality(model_fit = stan_output_combined ,model_data = stan_data, cov_name = 'rho',
               wd_values = c(200,800), growth_name = 'growth_dt', mortality_curve = TRUE, xaxis_on = TRUE, legend = FALSE)
```

**Fig 8** Predicted hazard rates (top) and mortality probability (bottom) for an average 200 (black) and 800 kg/m2 (red) kg/m2 wood density species. Errors are 95% credible intervals. rho = wood density.

```{r 3d plots, fig.height=10,fig.width=10, echo=FALSE}
par(mfrow=c(2,1))
plot3d(model_fit = stan_output_combined,
        model_data = stan_data, cov_name = 'rho',
        wd_values = seq(200,800, by=10),
        theta=145, phi=20,mortality_curve = FALSE)

plot3d(model_fit = stan_output_combined,
       model_data = stan_data, cov_name = 'rho',
       wd_values = seq(200,800, by=10),
       theta=145, phi=20,mortality_curve = TRUE)
```

**Fig 9** Mean predicted 3D surface for hazard rates (top) and mortality probability (bottom). Note that there is no 'significant' difference at low growth rate between low and high wood density. rho = wood density,This can be better seen in Fig. 8.

##Notes/Questions

@@

## K-fold Cross validation example excerp from [Vehtari & Gelman 2014](http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf)

To implement K-fold cross-validation in Stan we need to repeatedly fit the model to one subset of the data and use it to predict the held-out set. The way we recommend setting this up is to do the partitioning in R (or Python, or whatever data-processing environment is being used) and then pass the training data and held-out data to Stan in two pieces.
With linear regression, for example, we start with the Stan model on page 4, passing in the training data as N, y, X and the held-out set as N holdout, y holdout, X holdout, with the data block augmented accordingly. We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data. The code looks like this:

```{r eval=FALSE}
􏰋
￼￼￼data { int N;
  int J;
vector[N] y;
matrix[N,J] X;
int N_holdout;
vector[N_holdout] y_holdout;
matrix[N_holdout,J] X_holdout;
}
  parameters {
    vector[J] b;
real<lower=0> sigma;
  }
  model {
    y ~ normal(X*b, sigma);
increment_log_prob(-log(sigma));
  }
  generated quantities {
    vector[N_holdout] log_lik;
for (n in 1:N_holdout){
  // data model
// log prior for p(sigma) propto 1/sigma
log_lik[n] <- normal_log(y_holdout[n], X_holdout[n]*b, sigma);
9
} }

@@

##References

>>>>

## Introduction

Understanding who lives or dies, and why, is critical to accurately predict population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr]. In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more likely to die [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr]. Empirical studies have broadly supported this assumption, with many studies showing that the probability of dying decreasing exponentially with growth rate [Fig 1; @Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw], and consequently most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db] and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use this growth-mortality relationship as the predominant, and in many cases, only cause of plant mortality [@Hawkes:2000ib; @McDowell:2011dr].

@@

## Data

@@

## Model Description

@@

## Analysis plan

Step 1 - define a general structure of model we think will capture data, on basis of prior knowledge
Step 2 - assuming species effects on all pars, ask which growth meausre best predicts data, compare 4 growth rates using cross validation
Step 3 - fit full model with trait effects, look at effect sizes of WD on a, b and c
Step 4 - based on 3, choose some subsets of models, compare these using both bayes factor (most parsimonious) and predictive capacity

@@

## Results
**Take home message:**
Wood density decreases intercept (log_a) & further increases the steepness of the growth-mortality slope (log_b). It also decreases the growth-independent hazard log_ho
\pagebreak

## Notes/Questions

@@

## Cross validation

We will implement *K-fold cross validation*.

According to [Vehtari & Gelman 2014](http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf):

  To implement K-fold cross-validation in Stan we need to repeatedly fit the model to one subset of the data and use it to predict the held-out set. The way we recommend setting this up is to do the partitioning in R (or Python, or whatever data-processing environment is being used) and then pass the training data and held-out data to Stan in two pieces. With linear regression, for example, we start with the Stan model on page 4, passing in the training data as N, y, X and the held-out set as N holdout, y holdout, X holdout, with the data block augmented accordingly. We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data.

For a model $M$, training data $D$, and test data $D^*$, we want to calculate

$$p(D^* | D,M) = \int p(\theta | D, M) \, p(D^* | \theta, M) \, d \theta$$

The distribution $p(\theta | D, M)$ gives the likelihood of parameter values $\theta$ given the training data (obtained from fitting the model) and is calculated (via Bayes Rule) as

$$p(\theta | D, M) = \frac{p(\theta) \, p(D|\theta)}{p(D)}.$$

$p(D^* | \theta, M)$ is then the probability of observing the test data given a particular parameter set (i.e the likelihood of the data) and is calculated as
$$p(D^* | \theta, M) = \prod_i p(D^*_i | \theta, M).$$

Log-likelihoods are easier to deal with and so we have

$$\log p(D^* | \theta, M) = \sum_i \log p(D^*_i | \theta, M).$$

Aki \& Gelman 2014 provide methods for estimating "an importance-sampling approximated LOO directly using the log-likelihood evaluated at the posterior simulations of the parameter values"

@@

## References
